Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2021-07-05 source/sec_SGD.tex
Update on Overleaf.

We first show that FedAvg has an $\mathcal{O}(1/KT)$ convergence rate
for $\mu$-strongly convex and $L$-smooth objectives. The result relies on a technical improvement over the analysis in~\cite{li2019convergence}. Moreover, it
implies a distinction in communication efficiency that guarantees
this linear speedup for FedAvg with full and partial device participation.
With full participation, $E$ can be chosen as large as $\mathcal{O}(\sqrt{T/N})$
without degrading the linear speedup in the number of workers. On
the other hand, with partial participation, $E$ must be $\mathcal{O}(1)$
to guarantee $\mathcal{O}(1/KT)$ convergence.
\begin{theorem}
	\label{thm:SGD_scvx}Let $\overline{\mathbf{w}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{w}_{T}^{k}$ in FedAvg,
	$\nu_{\max}=\max_{k}Np_{k}$, and set decaying learning rates $\alpha_{t}=\frac{4}{\mu(\gamma+t)}$
	with $\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then
	under Assumptions~\ref{ass:lsmooth} to \ref{ass:subgrad2} with full device participation, 
	% under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with at most $K$ sampled devices
	at each communication round, 
	{\small \begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}}
	\label{th:scvx_sgd}
\end{theorem}
\textbf{Outline of proof.} Because our unified analyses of results in the main text follow the same framework with variations in technical details, we first give an outline of proof for Theorem~\ref{th:scvx_sgd} to illustrate the main ideas and highlight the improvement of our analyses over prior works.\\
\textit{One-step progress bound.} The analysis starts with the full participation setting, and establishes a recursive contraction bound for the distance to the optimal solution after one step of the FedAvg algorithm:
\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq (1-\alpha_t)\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}+ \cO(\alpha_t^2 \sigma^2/N + \alpha_t^3E^2G^2).
\end{align*}
% \lkxcom{explain variance here and the improvement?}
The above bound consists of three components: the distance to the optimal solution 
in the previous step, 
the variance of stochastic gradients in local clients, and the variance
across different clients. This last term $\mathcal{O}(\alpha_{t}^{3}E^{2}G^{2})$ is the key improvement over ~\cite{li2019convergence}, which has $\mathcal{O}(\alpha_{t}^{2}E^2 G^2)$ instead.\\
\textit{Iterative deduction.} This step applies induction to the recursive one-step progress bounds to obtain a non-recursive bound on $\mathbb{E}\|\ov{w}_{T}-\vw^{\ast}\|^{2}$ in terms of the initial distance ($\|\ov{w}_{0}-\vw^{\ast}\|^{2}$), which is then converted to a bound on $\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}$ using $L$-smoothness.\\
\textit{From full to partial participation.} In addition to the variances arising from stochastic updates at local clients and averaging across clients, partial participation, which involves a sampling procedure, adds a an additional term $\mathcal{O}(\frac{1}{K} \alpha_t^2 E^2G^2)$ of leading order resulting from the sampling variance is added to the recursive contraction bound.

>>>>

We first show that FedAvg has an $\mathcal{O}(1/KT)$ convergence rate
for $\mu$-strongly convex and $L$-smooth objectives. The result relies on a technical improvement over the analysis in~\cite{li2019convergence}. Moreover, it
implies a distinction in communication efficiency that guarantees
this linear speedup for FedAvg with full and partial device participation.
With full participation, $E$ can be chosen as large as $\mathcal{O}(\sqrt{T/N})$
without degrading the linear speedup in the number of workers. On
the other hand, with partial participation, $E$ must be $\mathcal{O}(1)$
to guarantee $\mathcal{O}(1/KT)$ convergence.
\begin{theorem}
	\label{thm:SGD_scvx}Let $\overline{\mathbf{w}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{w}_{T}^{k}$ in FedAvg,
	$\nu_{\max}=\max_{k}Np_{k}$, and set decaying learning rates $\alpha_{t}=\frac{4}{\mu(\gamma+t)}$
	with $\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then
	under Assumptions~\ref{ass:lsmooth} to \ref{ass:subgrad2} with full device participation, 
	% under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with at most $K$ sampled devices
	at each communication round, 
	{\small \begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}}
	\label{th:scvx_sgd}
\end{theorem}
\textbf{Outline of proof.} Because our unified analyses of results in the main text follow the same framework with variations in technical details, we first give an outline of proof for Theorem~\ref{th:scvx_sgd} to illustrate the main ideas and highlight the improvement of our analyses over prior works.\\
\textit{One-step progress bound.} The analysis starts with the full participation setting, and establishes a recursive contraction bound for the distance to the optimal solution after one step of the FedAvg algorithm:
\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq (1-\alpha_t)\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}+ \cO(\alpha_t^2 \sigma^2/N + \alpha_t^3E^2G^2).
\end{align*}
% \lkxcom{explain variance here and the improvement?}
The above bound consists of three components: the distance to the optimal solution 
in the previous step, 
the variance of stochastic gradients in local clients, and the variance
across different clients. This last term $\mathcal{O}(\alpha_{t}^{3}E^{2}G^{2})$ is the key improvement over ~\cite{li2019convergence}, which has $\mathcal{O}(\alpha_{t}^{2}E^2 G^2)$ instead.\\
\textit{Iterative deduction.} This step applies induction to the recursive one-step progress bounds to obtain a non-recursive bound on $\mathbb{E}\|\ov{w}_{T}-\vw^{\ast}\|^{2}$ in terms of the initial distance ($\|\ov{w}_{0}-\vw^{\ast}\|^{2}$), which is then converted to a bound on $\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}$ using $L$-smoothness.\\
\textit{From full to partial participation.} In addition to the variances arising from stochastic updates at local clients and averaging across clients, partial participation, which involves a sampling procedure, gives rise to a third variance term. adds a an additional term $\mathcal{O}(\frac{1}{K} \alpha_t^2 E^2G^2)$ of leading order resulting from the sampling variance is added to the recursive contraction bound.
